A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-06 introduction.tex
SMT-LIB example

To solve \cref{ex:string-constraints} using \Calculus{},
\cref{const:x-in-c-dd,const:y-in-b,const:x-substring} can be translated to the
product of Parikh automata $B \times A'$, where $A'$ (\cref{fig:aut_a}) is the
Parikh automata obtained by applying the encoding of $\Substring$ described in
\cite{ostrich-plus}. That description can then be used to obtain values for
$\CountOf{\Char{a}}(y)$ and $\CountOf{\Char{b}}(y)$. Here follows an intuition
for our method on $B \times A'$ under the global constraint
\cref{const:more-a-than-b}. Note that the cost registers of \cref{fig:aut_b} are
simply counting letters, while the extra registers $l, r$ in $A'$ also keeps
track of the matched substring. However, these registers can be mostly
disregarded here, since we have no constraints on their values.

Our approach to the computation of the Parikh image (and therefore the solution
to the intersection problem posed by \cref{ex:string-constraints}) consists of
two applications of laziness; late computation of products of automata, and lazy
enforcement of connectivity constraints. Both are in contrast to the eager
approach to finding the Parikh image described in~\cite{generate-parikh-image},
that would have started by computing the product $A' \times B$, translated it to
a Presburger formulae with $\CountOf{\Char{a}}$ etc as free variables, and then
plugged in the constraint \cref{const:more-a-than-b}.

@@

Performing symbolic reasoning on these equations and plugging the corresponding
symbolic representation of the constraints on each corresponding integer
variable back as an annotation on its transition, we get obtain the two automata
in \cref{fig:propagated}.

Several transitions are already marked as unuseable (have a transition variable
value of $0$). This includes any transition for the letter $\Char{d}$, as
$\CountOf{\Char{d}}(B) = \sum_{\Transition} 0 = 0$, since the register for
$\Char{d}$ is zero for all transitions in $B$. By removing such transitions to
obtain new automata and computing their product, we obtain the product automaton
of \cref{fig:product}.

@@

By putting off computing the product $B \times A'$ until after performing linear
reasoning on the number of times each transition must be taken and using that to
prune the terms, we have computed a smaller product than we would have with a
na\"ive approach. Moreover, since this automaton contains only one loop along
one main path, all runs through it can be described by the flow-preservation
equations described above without the need for costly connectivity-enforcing
constraints as in~\cite{generate-parikh-image}. This means that we can now fully
represent all possible solutions to \cref{ex:string-constraints} by the linear
equation $\CountOf{\Char{a}} = 2 \land \CountOf{\Char{c}} > 0 \land
\CountOf{\Char{b}} = 0 \land \CountOf{\Char{d}} = 0$, derived from applying the
same flow encoding as before.

>>>>

To solve \cref{ex:string-constraints} using \Calculus{},
\cref{const:x-in-c-dd,const:y-in-b,const:x-substring} can be translated to the
product of Parikh automata $B \times A'$, where $A'$ (\cref{fig:aut_a}) is the
Parikh automata obtained by applying the encoding of $\Substring$ described in
\cite{ostrich-plus}. That description can then be used to obtain values for the
subsstring offsets $i, n$. Here follows an intuition for our method on $B \times
A'$ under the global constraint on the substring \cref{const:more-a-than-b}.

Our approach to the computation of the Parikh image (and therefore the solution
to the intersection problem posed by \cref{ex:string-constraints}) consists of
two applications of laziness; late computation of products of automata, and lazy
enforcement of connectivity constraints. Both are in contrast to the eager
approach to finding the Parikh image described in~\cite{generate-parikh-image},
that would have started by computing the product $A' \times B$, translated it to
a Presburger formula with $i, n$ etc as free variables, and then plugged in the
constraint \cref{const:more-a-than-b}.

@@

Performing symbolic reasoning on these equations and plugging the corresponding
symbolic representation of the constraints on each corresponding integer
variable back as an annotation on its transition, we obtain the two automata in
\cref{fig:propagated}.

% FIXME B-loopen inte nollad här! Uppdatera!

Several transitions are already marked as unuseable (have a transition variable
value of $0$). This includes any transition for the letter $\Char{d}$, as
$\CountOf{\Char{d}}(B) = \sum_{\Transition} 0 = 0$, since the register for
$\Char{d}$ is zero for all transitions in $B$.

Note that the self-loop on state $B$ is nonzero due to the above mentioned
self-balancing. However, a reachability computation will conclude that the state
is unreachable, and allow us to conclude that all its outgoing transitions must
be zero.

After removing all transitions concluded to be unuseable to produce new,
equivalent automata, we compute the product in \cref{fig:product}.

@@

By putting off computing the product $B \times A'$ until after performing linear
reasoning on the number of times each transition must be taken and using that to
prune the terms, we have computed a smaller product than we would have with a
na\"ive approach. Moreover, since this automaton contains only one loop along
one main path, all runs through it can be described by the flow-preservation
equations described above without the need for costly connectivity-enforcing
constraints as in~\cite{generate-parikh-image}. We can now obtain a generalised
solution to our constraints by inspection.

### 2023-06-06 main.tex
SMT-LIB example

%% Appendix
\appendix
\section{Appendix}
\subsection{Per-instance comparison between backends}
\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:baseline}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and baseline]{The plot shows most dots at the bottom of the plot, with some plots collected at the right-hand side.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:nuxmv}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and \Nuxmv{}]{The plot shows flames of dots between the bottom and top, with a line of dots near the right-hand side corresponding to timeouts.}
  \end{subfigure}
\end{figure}
\subsection{Example instance for \Catra{}}
\begin{lstlisting}[caption={An example input file for \Catra{} for \cref{ex:string-constraints} in \cref{sec:intuition}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), labels (except those with ranges), register increments, and constraints on the final values of their counters.}, label=lst:input-example]
\input{example.par}
\end{lstlisting}

% Text of appendix \ldots

>>>>

%% Appendix
\appendix
\section{Appendix}
\subsection{Per-instance comparison between backends}
\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:baseline}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and baseline]{The plot shows most dots at the bottom of the plot, with some plots collected at the right-hand side.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter.pdf}
    \caption{Per-instance runtime between baseline and \Calculus{}}
    \label{fig:duels:nuxmv}
    \Description[A scatter plot showing individual instances as dots compared to \Calculus{} and \Nuxmv{}]{The plot shows flames of dots between the bottom and top, with a line of dots near the right-hand side corresponding to timeouts.}
  \end{subfigure}
\end{figure}

\subsection{Example instance for \Catra{}}
\lstinputlisting[caption={An example input file for \Catra{} for \cref{ex:string-constraints} in \cref{sec:intuition}, illustrating every major syntax element. From beginning to end: synchronised (product) automata using the keyword \texttt{synchronised} (automata A and B), labels (except those with ranges), register increments, and constraints on the final values of their counters.}, label=lst:input-example]{example.par}

\subsection{SMT-LIB encoding of the string constraint example}
\lstinputlisting[caption={An SMT-LIB encoding of \cref{ex:string-constraints}.}, label=lst:string-constraints]{example.smt2}
